Competing Bandits:\\ The Perils of Exploration under Competition \thanks{This is a merged and final version of two conference papers, \citet{CompetingBandits-itcs18} and \citet{CompetingBandits-ec19}, with a unified and streamlined presentation.}

### 2020-07-09 journal paper/content/sec-intro.tex
Added a sentence.
Minor edits in "econ interpretation: network effects".

Our decision rules differ in terms of rationality: from fully rational decisions with \HardMax to relaxed rationality with \HardMaxRandom to an even more relaxed rationality with \SoftMaxRandom. The same distinctions also control the severity of competition between the principals: from cut-throat competition with \HardMax to a more relaxed competition with \HardMaxRandom, to an even more relaxed competition with \SoftMaxRandom. Indeed, with \HardMax you lose all customers as soon as you fall behind in performance, with \HardMaxRandom you get some small market share no matter what, and with \SoftMaxRandom you are further guaranteed a market share close to $\tfrac12$ as long as your performance is not much worse than the competition. The uniform choice among principals corresponds to no rationality and no competition.

@@

\item We investigate the first-mover advantage as a different channel to vary the intensity of competition: from the first-mover to simultaneous arrival to late-arriver. (We focus on the \HardMax decision rule.) We find that the first-mover is incentivized to choose a more advanced exploration algorithm, whereas the late-arriver is often incentivized to choose the ``greedy algorithm" (more so than under simultaneous arrival). Consumer welfare is higher under early/late arrival than under simultaneous entry. We frame these results in terms inverted-U relationships.

@@

\ascomment{Guy: "anti-trust xhdr" synthesises several points from various places, pls see if I got it right.}

\xhdr{Economic interpretation: data network effects.}
\gaedit{Our model speaks to the ongoing debate on the role that data plays in the digital economy. One fundamental question in this debate is whether data can serve a similar role as traditional network effects, whereby competition for the market is more important than competition in the market due to the winner-take-all nature of such markets \citep{Rysman09, jullien2019economics}. The ``death spiral" effect that we find has a similar flavor as this as once a firm engages in exploration it gets completely starved of users. Indeed, we further find that a small data advantage for one firm gets amplified under competition and leads to that firm taking the entire market. A distinctive feature of our approach is that we explicitly model the learning problem of the firms and consider them as deploying algorithms for solving this problem.  Thus, we do not explicitly model such network effects, but rather they arise endogenously from our setup.

Our results highlight that understanding the performance of learning algorithms in isolation does not necessarily translate to understanding their impact in competition, precisely due to the fact that competition leads to the endogenous generation of data observed by the firms. Approaches such as \cite{bajari2018impact} that argue that the diminishing returns to scale and scope of data in isolation render such data feedback loops as non-existent ignore the differences induced by learning in isolation versus under competition. Furthermore, explicitly incorporating the interaction between learning technology and data creation allows us to speak on how data advantages are characterized and amplified not only by data quantity, but also the increased data quality gathered by better learning algorithms.
}

@@

Our study has a dual purpose: (i) shed light on real-world implications of some typical scenarios, and (ii) investigate the space of models for describing the real world. As an example to clarify the latter point, consider the \HardMax model with simultaneous entry. It is is not necessarily the most realistic model, but arguably the most natural one to study \emph{a priori}. However, our results elucidate the need for more refined models with ``free exploration" (\eg via random agents or early entry).

>>>>

Our decision rules differ in terms of rationality: from fully rational decisions with \HardMax to relaxed rationality with \HardMaxRandom to an even more relaxed rationality with \SoftMaxRandom. The same distinctions also control the severity of competition between the principals: from cut-throat competition with \HardMax to a more relaxed competition with \HardMaxRandom, to an even more relaxed competition with \SoftMaxRandom. Indeed, with \HardMax you lose all customers as soon as you fall behind in performance, with \HardMaxRandom you get some small market share no matter what, and with \SoftMaxRandom you are further guaranteed a market share close to $\tfrac12$ as long as your performance is not much worse than the competition. The uniform choice among principals corresponds to no rationality and no competition. \asedit{While agents' rationality and severity of competition are often modeled separately in the literature, it is not unusual to have them modeled with the same ``knob" \cite[\eg][]{Gabaix-16}.}

@@

\item We investigate the first-mover advantage as a different channel to vary the intensity of competition: from the first-mover to simultaneous arrival to late-arriver. (We focus on the \HardMax decision rule.) We find that the first-mover is incentivized to choose a more advanced exploration algorithm, whereas the late-arriver is often incentivized to choose the ``greedy algorithm" (more so than under simultaneous arrival). Consumer welfare is higher under early/late arrival than under simultaneous entry. We frame these results in terms of an inverted-U relationship.

@@

\ascomment{The xhdr below is basically from Guy, with minor edits by Alex.}

\xhdr{Economic interpretation: network effects \asedit{of data}.}
Our model speaks to the ongoing debate on the role that data plays in the digital economy. One fundamental question in this debate is whether data can serve a similar role as traditional network effects, whereby competition \emph{for} the market is more important than competition \emph{in} the market, due to the winner-take-all nature of \asedit{the digital economy} \citep{Rysman09, jullien2019economics}. 
The ``death spiral" effect that we find has a similar flavor: 
\asdelete{as this as}
once a firm engages in exploration it gets completely starved of users. We further find that a small data advantage for one firm gets amplified under competition and leads to that firm taking the entire market. A distinctive feature of our approach is that we explicitly model the learning problem of the firms and consider them deploying algorithms for solving this problem.  Thus, we do not explicitly model \asedit{the} network effects, but they arise endogenously from our setup.

\ascomment{Guy: could you pls clarify the diff between "for the market" and "in the market". Is this the distinction between "competing for market share" and "competing for profit"? If so, I understand what you mean by "for" the market; but why "in" the market? Perhaps a small edit would make it much clearer?}

Our results highlight that understanding the performance of learning algorithms in isolation does not necessarily translate to understanding their impact in competition, precisely due to the fact that competition leads to the endogenous generation of data observed by the firms. Approaches such as \citet{bajari2018impact} argue that the diminishing returns to scale and scope of data in isolation \asedit{mitigate} such data feedback loops,
%as non-existent
\asedit{but} ignore the differences induced by learning in isolation versus under competition. Furthermore, explicitly incorporating the interaction between learning technology and data creation allows us to speak on how data advantages are characterized and amplified not only by data quantity, but also the increased data quality gathered by better learning algorithms.

\ascomment{Guy: Previously you also cited \cite{varian2018artificial,lambrecht2015can} for the same thing as \citet{bajari2018impact}. Why not now? Also, previously you explicitly mentioned "barriers to entry" and "market tipping" and "anti-trust implications". Why not now? I don't really know any of this literature, so I defer to your judgement on this ...}

@@

Our study has a dual purpose: (i) shed light on real-world implications of some typical scenarios, and (ii) investigate the space of models for describing the real world. As an example to clarify the latter point, consider the \HardMax model with simultaneous entry. It is is not necessarily the most realistic model, but arguably the most natural one to study \emph{a priori}. However, our results elucidate the need for more refined models that allow for ``free exploration" (\eg via random agents or early entry).
